---
title: "1a_primates"
author: "Jenny Yijian Huang"
date: '2022-08-01'
output: html_document
---

# -------- TO DO --------- #

# Set the directories below to correspond to paths on your machine:
```{r}
# The directory where the analysis is performed:
wd_path <- '/Users/JennyH/Desktop/PSR_Model/'
# The directory where the original data are:
data_path <- 'Data/'
# The directory where the processed data should be saved:
save_path <- 'Data/'
# Whether the processed data should be saved or not:
save_files <- TRUE
```

```{r}
library(tidyverse)
library(devtools)

devtools::install_github("gpapadog/BiasedNetwork")
```

# --------- BEGINNING -------- #

```{r}
# Setting the working directory.
setwd(wd_path)

# Loading libraries.
library(data.table)
library(gplots)
```

# ---------- PART A: Loading in the data -------------- #

# ---------- PART A and B: Loading in the data -------------- #

```{r}
parasite_traits_original <- read.csv("parasite_list.csv")
head(parasite_traits_original)
primate_traits_original <- read.csv("trait_parasite_species_cleaned.csv") # trophic category, group size, homerange.
head(primate_traits_original)
host_parasite_interactions <- read.csv("GMPD_unique_parasite_sp.csv")
head(host_parasite_interactions)
```

```{r}
primate_traits_original <- primate_traits_original %>% 
  rename(social_time = Mean_Percent_Social_Activity_Time)
```


### EDA: feature distributions.
```{r}
hist(primate_traits_original$bodymass_kg)
hist(primate_traits_original$social_time)
hist(log(primate_traits_original$mean_groupsize)) # log
hist(log(primate_traits_original$mean_dayrange_meters)) # log
hist(log(primate_traits_original$homerange_ha)) # log

ggplot(primate_traits_original,aes(mating_system,mating_system))+
  geom_bar(stat="identity",width=1)+
  theme(axis.text.x = element_text(angle = 90))
ggplot(primate_traits_original,aes(realm,realm))+
  geom_bar(stat="identity",width=1)+
  theme(axis.text.x = element_text(angle = 90))
ggplot(primate_traits_original,aes(trophic_folivore,trophic_folivore))+
  geom_bar(stat="identity",width=1)+
  theme(axis.text.x = element_text(angle = 90))
ggplot(primate_traits_original,aes(trophic_Omnivore,trophic_Omnivore))+
  geom_bar(stat="identity",width=1)+
  theme(axis.text.x = element_text(angle = 90))
ggplot(primate_traits_original,aes(iucn,iucn))+
  geom_bar(stat="identity",width=1)+
  theme(axis.text.x = element_text(angle = 90))
```



```{r}

# View(host_parasite_interactions)
# clean original df to select variables of interest
primate_traits <- primate_traits_original %>% 
  select(HostCorrectedName, bodymass_kg, SSD,
         social_time, mean_groupsize, mean_dayrange_meters, homerange_ha,
         mating_system, trophic_folivore, trophic_Omnivore, realm, iucn,
         diel_activity, habitat_category, locomotion)

parasite_traits <- parasite_traits_original %>% 
  rename(ParasiteCorrectedName = parasite) %>% 
  select(-X)

HP_interactions <- host_parasite_interactions %>% 
  rename(HostCorrectedName = HostCorrectedName_Corbet) %>% 
  select(ParasiteCorrectedName, HostCorrectedName, Citation, LocationName)

# perform a 3 way join of tables to create traits_and_studies, a df of 
# all observed interactions and the desired traits of the host and the parasite.
a1 <- merge(HP_interactions, primate_traits, by = "HostCorrectedName")
traits_and_studies <- merge(a1, parasite_traits, by="ParasiteCorrectedName")


# Keeping track of the unique primates and parasite species in the study and their
# number. Also keeping track of the unique studies.
uni_primates <- unique(traits_and_studies$HostCorrectedName)
uni_parasites <- unique(traits_and_studies$ParasiteCorrectedName)
uni_studies_lst <- unique(traits_and_studies$Citation)
nM <- length(uni_primates)
nP <- length(uni_parasites)
nS <- length(uni_studies_lst)
cat(nM, nP, nS) # 93 157 217
```


# Do the ‘bipartitedf’ entries make sense:
```{r}
traits_and_studies %>% 
  filter(#HostCorrectedName == "Varecia_variegata")
         ParasiteCorrectedName == "Shigella sp.")
```

# ------ PART C: Array of observed interactions -------- #

# Creating an array with dimensions that correspond to birds, plants and
# studies. Entries are equal to 0 if the species did not interact in the
# specific study, and equal to 1 if they were recorded to interact.

```{r}
obs_A2 <- array(0, dim = c(nM, nP, nS))
dimnames(obs_A2) <- list(uni_primates, uni_parasites, uni_studies_lst)
for (ss in 1 : nrow(traits_and_studies)) {
  wh1 <- which(uni_primates == traits_and_studies$HostCorrectedName[ss])
  wh2 <- which(uni_parasites == traits_and_studies$ParasiteCorrectedName[ss])
  wh3 <- which(uni_studies_lst == traits_and_studies$Citation[ss])
  obs_A2[wh1, wh2, wh3] <- 1
}

dim(obs_A2) # primate, parasite, studies 

sum(obs_A2) # 732 ones.
sum(obs_A2) / (nM*nP) # 5.01 % of adjacency matrix is ones.

if (save_files) {
  save(obs_A2, file = paste0(save_path, 'obs_A2.dat'))
}
```

# ------ PART D: Covariates and correlation matrices -------- #

# Here we re-format the original data to acquire matrices of covariate
# information for each set of species, and the phylogenetic correlation
# matrices.

# Starting for the first set of species.

# Including the bird covariate information in one data frame. This data frame
# includes phylogenetic information that will be used to create the species'
# correlation matrix.
#
```{r}
all_X2 <- traits_and_studies %>% 
  select(HostCorrectedName,
         bodymass_kg,
         mean_groupsize, 
         mean_dayrange_meters,
         homerange_ha,
         social_time,
         SSD,
         diel_activity,
         habitat_category,
         locomotion,
         mating_system,
         trophic_folivore,
         trophic_Omnivore,
         realm,
         iucn)

# Excluding replicates of primates:
all_X2 <- unique(all_X2)
# Ensuring that the dimensions are correct:
cat(nM, dim(all_X2)[1])
# Ensuring that the species in all_X are in the same order as in obs_A:
sum(all_X2$HostCorrectedName != dimnames(obs_A2)[[1]])  # Should be equal to 0.
```

# Some information on the covariates:
body mass, ssd, group size, homerange, social time, nocturnal, arboreal forest, mating-system, diet, realm, iucn
```{r}
pM <- c(6, 13)  # Two continous and three binary covariates:
obs_X2 <- matrix(NA, nrow = nM, ncol = sum(pM))
rownames(obs_X2) <- uni_primates
colnames(obs_X2) <- c('logBodyMass', 'logSSD', 'log_mean_groupsize',
                      'log_mean_dayrange_meters', 'loghomerange_ha', 
                      'social_time', 'isnocturnal', 'isarboreal', 
                      'isforest', 'ismultimale-multifemale', 'ismonogamous', 
                      'isdispersed','isfolivorous','isomnivorous',
                      'realm_M_Africa','realm_Neotropics','realm_Asia', 
                      'iucn_endangered', 'iucn_least_concern')
obs_X2[, 1] <- log(all_X2$bodymass_kg) # taking logs of body mass
obs_X2[, 2] <- log(all_X2$SSD)

obs_X2[, 3] <- log(all_X2$mean_groupsize)
obs_X2[, 4] <- log(all_X2$mean_dayrange_meters)
obs_X2[, 5] <- log(all_X2$homerange_ha)
obs_X2[, 6] <- all_X2$social_time

obs_X2[, 7] <- (all_X2$diel_activity == 'nocturnal') * 1
obs_X2[, 8] <- (all_X2$locomotion == 'arboreal') * 1
obs_X2[, 9] <- (all_X2$habitat_category == 'forest') * 1

obs_X2[, 10] <- (all_X2$mating_system == 'multimale-multifemale') * 1
obs_X2[, 11] <- (all_X2$mating_system == 'monogamy') * 1
obs_X2[, 12] <- (all_X2$mating_system == 'dispersed') * 1

obs_X2[, 13] <- (all_X2$trophic_folivore == 'Folivore') * 1
obs_X2[, 14] <- (all_X2$trophic_Omnivore == 'Omnivore/Insectivore') * 1

obs_X2[, 15] <- (all_X2$realm == 'M_Africa') * 1
obs_X2[, 16] <- (all_X2$realm == 'Neotropics') * 1
obs_X2[, 17] <- (all_X2$realm == 'Asia') * 1

obs_X2[, 18] <- (all_X2$iucn == 'endangered') * 1
obs_X2[, 19] <- (all_X2$iucn == 'least concern') * 1

if (save_files) {
  save(obs_X2, file = paste0(save_path, 'obs_X2.dat'))
}

```


# For the second set of species: We perform a similar process to acquire their phylogenetic correlation matrix and their covariate information.

```{r}
all_W2 <- traits_and_studies %>% 
  select(ParasiteCorrectedName,
         parasite_type,
         transmission_type,
         life_cycle 
         )

# Excluding species that are duplicated:
all_W2 <- unique(all_W2)
####
# Debug1
y <- rnorm(157, mean = 3, sd = 1.0)
all_W2 <- cbind(all_W2, y)
####

# Making sure that the dimension and order of species is correct:
cat(nP, dim(all_W2)[1])
sum(all_W2$Parasite_Species != dimnames(obs_A2)[[2]])  # Should be 0.
```
# VECTOR OF 1s AS PLACEHOLDER FOR A CONTINUOUS FEATURE.
```{r}
y <- rnorm(nP,1,0)
```

```{r}
pP <- c(1, 11)  # Number of continuous and binary covariates.
obs_W2 <- matrix(NA, nrow = nP, ncol = sum(pP))

obs_W2[, 1] <- log(all_W2$y) # VECTOR OF 1 AS PLACEHOLDER FOR A CONTINUOUS FEATURE.
obs_W2[, 2] <- (all_W2$parasite_type == 'Bacteria') * 1
obs_W2[, 3] <- (all_W2$parasite_type == 'Virus') * 1
obs_W2[, 4] <- (all_W2$parasite_type == 'Arthropod') * 1
obs_W2[, 5] <- (all_W2$parasite_type == 'Protozoa') * 1
obs_W2[, 6] <- (all_W2$transmission_type == 'CloseT') * 1
obs_W2[, 7] <- (all_W2$transmission_type == 'Nonclose') * 1
obs_W2[, 8] <- (all_W2$transmission_type == 'Vector') * 1
obs_W2[, 9] <- (all_W2$transmission_type == 'Intermediate') * 1
obs_W2[, 10] <- (all_W2$transmission_type == 'Sexual_Vertical_CloseT') * 1
obs_W2[, 11] <- (all_W2$life_cycle == "Direct") * 1
obs_W2[, 12] <- (all_W2$parasite_type == 'Helminth') * 1

# Creating the column names.
colnames(obs_W2) <- c('isHelminth', 'isBacteria', 'isVirus', 'isArthropod',
                     'isProtozoa', 'isCloseT', 'isNonclose', 'isVector',
                     'isIntermediate', 'isSexualVerticalCloseT', 'isDirect', 'y')

if (save_files) {
  save(obs_W2, file = paste0(save_path, 'obs_W2.dat'))
}
```

# ------ PART E: Focus and species occurrence -------- #

Are there studies in the GMPD in which multiple host species are recorded?
```{r}
# For occurrence of network study
obs_OB2 <- array(0, dim = c(nM, nS))
obs_OP2 <- array(0, dim = c(nP, nS))
dimnames(obs_OB2) <- list(uni_primates, uni_studies_lst)
dimnames(obs_OP2) <- list(uni_parasites, uni_studies_lst)

studied_multiple_hosts <- array(0, nS)
for (ss in 1 : nS) {
  wh_study <- uni_studies_lst[ss]
  these_primates <- unique(traits_and_studies$HostCorrectedName[traits_and_studies$Citation == wh_study]) # primates that have been studies here.
  if(length(these_primates) > 1){
    studied_multiple_hosts[ss] = 1
  }
}

sum(studied_multiple_hosts) # 49
```

```{r}
multihost_studies <- traits_and_studies %>% 
  group_by(Citation) %>% 
  select(Citation, HostCorrectedName) %>% 
  unique() %>% 
  group_by(Citation) %>% 
  arrange(Citation) %>% 
  filter(n()>1) %>% # citation appears several times, so these are the studies with multiple hosts.
  distinct(Citation) %>% 
  select(Citation) # this turns the df into a list of citations.
```
Out of our 217 studies, we have 49 that study multiple hosts, although some study only one genera.

```{r}
multibug_studies <- traits_and_studies %>% 
  group_by(Citation, ParasiteCorrectedName) %>% 
  select(ParasiteCorrectedName) %>% 
  unique() %>% # 
  group_by(Citation) %>% 
  arrange(Citation) %>% 
  filter(n()>1) %>% # those citations that studied more that one parasite.
  distinct(Citation) %>% 
  select(Citation) # this turns the df into a list of citations.
```
Out of our 217 studies, we have 49 that study multiple hosts, 108 that study multiple parasites, and 38 that study both multiple hosts AND multiple parasites.

```{r}
network_studies <- intersect(multihost_studies, multibug_studies) %>% 
  distinct(Citation) %>% 
  select(Citation) # this turns the dataframe into a list of citations

# SANITY CHECK
"Petrasova et al. 2010" %in% network_studies$Citation
```
# ------ PART E: Focus and species occurrence across studies -------- #


# Creating two arrays that correspond to whether the studies would have
# recorded an observed interaction or not, and whether species co-occur in the
# study area. We will assume that these arrays are 0/1. 

# For the focus array, this makes sense as we have access to whether the study is animal or plant focused or a network study. 
# For the species occurrence, we know for a fact
# that a species with a recorded interaction occurs in the area. The only
# simplifying assumption is therefore when we set probability of occurrence to
# zero for species without any recorded interaction.
#

```{r}
# For study focus:
obs_F2 <- array(0, dim = c(nM, nP, nS))
dimnames(obs_F2) <- list(uni_primates, uni_parasites, uni_studies_lst)

for (ss in 1 : nS) {
    wh_study <- uni_studies_lst[ss]

    # i. Find all primates studied in the study.
    these_primates <- unique(traits_and_studies$HostCorrectedName[traits_and_studies$Citation == wh_study]) # the primates in the study
    
    # ii. find the indices of all 'these_parasites'
    wh_primates <- which(uni_primates %in% these_primates) # the indices of the primate.

    # iii. Find all parasite types studied in the study.
    wh_study_parasitetypesstudied <- unique(traits_and_studies$parasite_type[traits_and_studies$Citation == wh_study])
    
    # iv. For each parasite type, find all species in the database. We assume the test performed in the study has the potential to record such species.
    these_parasites <- unique(traits_and_studies$ParasiteCorrectedName[traits_and_studies$parasite_type %in% wh_study_parasitetypesstudied])
    
    # v. find the indices of all 'these_parasites'
    wh_parasites <- which(uni_parasites %in% these_parasites)
    
    obs_F2[wh_primates, wh_parasites, ss] <- 1 # records all of the potentially recorded parasites for the primates in the study.
}

if (save_files) {
  save(obs_F2, file = paste0(save_path, 'obs_F2.dat'))
}

obs_F2[,1,1] # all primates in row 1 column 1
obs_F2[1,,1] # all parasites in row 1 column 1
obs_F2[1,1,] # all studies in row 1 column 1
dim(obs_F2) # primate, parasite, study

sum(obs_F2) # 23456 ones.
sum(obs_F2) / (nM*nP*nS) #  0.007 % of matrix is ones.
```


```{r}
# For occurrence of network study
obs_OB2 <- array(0, dim = c(nM, nS))
obs_OP2 <- array(0, dim = c(nP, nS))
dimnames(obs_OB2) <- list(uni_primates, uni_studies_lst)
dimnames(obs_OP2) <- list(uni_parasites, uni_studies_lst)

for (ss in 1 : nS) {
  wh_study <- uni_studies_lst[ss]
  these_primates <- unique(traits_and_studies$HostCorrectedName[traits_and_studies$Citation == wh_study]) # primates that have been studied here.
  these_parasites <- unique(traits_and_studies$ParasiteCorrectedName[traits_and_studies$Citation == wh_study]) # primates that have been studied here.
  wh_primates <- which(uni_primates %in% these_primates)
  wh_parasites <- which(uni_parasites %in% these_parasites)
  obs_OB2[wh_primates, ss] <- 1 # indicator that the primate has been studied here.
  obs_OP2[wh_parasites, ss] <- 1 # indicator that the parasite has been studied here.
}

if (save_files) {
  save(obs_OB2, file = paste0(save_path, 'obs_OB2.dat'))
  save(obs_OP2, file = paste0(save_path, 'obs_OP2.dat'))
}

obs_OB2
```

